---
title: "Evaluate models"
output: html_document
date: "2025-08-21"
---

Loading R packages:
```{r}
library(tidyverse)
library(h2o)
library(pROC)
library(caret)
library(ROCR)
library(DALEX)
library(pdp)
library(ggsoccer)
library(RColorBrewer)
library(gridExtra)
library(h2o)
library(ingredients)
```

Loading data used to model:
```{r}
data_to_mod <- readRDS("data_to_mod.rds")
```

Setting testing dataset and features:
```{r}
h2o.init(nthreads = -1, max_mem_size = "4G")
data_h2o <- as.h2o(data_to_mod)
data_split = h2o.splitFrame(data = data_h2o, ratios = 0.8, seed=2025)
data_train = data_split[[1]]
data_test = data_split[[2]]

Y = "is_goal"
X2 = setdiff(h2o.names(data_train), Y)
X1 = c("distance_to_goal_line", "angle_to_goal")
```

Loading model with all features:
```{r}
modelname2 <- "" #Insert model name here
cmod2 <- h2o.loadModel(paste0("saved-models/", modelname2))
```

Variable importance:
```{r}
vi2 <- h2o.varimp(cmod2)
vi2
h2o.varimp_plot(cmod2, num_of_features = length(X2))
```

Model Performance:
```{r}
set.seed(2025)
pred <- h2o.predict(cmod2, newdata = data_test)
(perf <- h2o.performance(cmod2, newdata = data_test, valid=T))
```

Model Performance with another packages:
```{r}
data_test_r <- as.data.frame(data_test)
pred_r <- as.data.frame(pred)

roc <- roc(data_test_r$is_goal, pred_r$p1)
plot.roc(roc, print.auc = TRUE, print.thres = "best")

best_thres <- coords(roc, "best", ret = "threshold", transpose = T)
h2o.confusionMatrix(cmod2, data_test, thresholds = best_thres, metrics = "mean_per_class_accuracy")

pred_r$xG_bool <- ifelse(pred_r$p1 > best_thres, 1, 0)
(result <- confusionMatrix(data = factor(pred_r$xG_bool), 
                           reference = factor(data_test_r$is_goal), positive = "1", mode="prec_recall"))
```

Loading GLM model with all features:
```{r}
GLM_modelname2 <- "" #Insert model name here
GLM_cmod2 <- h2o.loadModel(paste0("saved-models/", GLM_modelname2))
```

GLM Variable importance:
```{r}
GLM_vi2 <- h2o.varimp(GLM_cmod2)
GLM_vi2
h2o.varimp_plot(GLM_cmod2, num_of_features = 13)
```

GLM Model Performance:
```{r}
set.seed(2025)
GLM_pred <- h2o.predict(GLM_cmod2, newdata = data_test)
(GLM_perf <- h2o.performance(GLM_cmod2, newdata = data_test, valid=T))
```

GLM Model Performance with another packages:
```{r}
GLM_data_test_r <- as.data.frame(data_test)
GLM_pred_r <- as.data.frame(GLM_pred)

GLM_roc <- roc(GLM_data_test_r$is_goal, GLM_pred_r$p1)
plot.roc(GLM_roc, print.auc = TRUE, print.thres = "best")

GLM_best_thres <- coords(GLM_roc, "best", ret = "threshold", transpose = T)
h2o.confusionMatrix(GLM_cmod2, data_test, thresholds = GLM_best_thres, metrics = "mean_per_class_accuracy")

GLM_pred_r$xG_bool <- ifelse(GLM_pred_r$p1 > GLM_best_thres, 1, 0)
(GLM_result <- confusionMatrix(data = factor(GLM_pred_r$xG_bool), 
                           reference = factor(GLM_data_test_r$is_goal), positive = "1", mode="prec_recall"))
```



Loading model with 2 features:
```{r}
modelname1 <- "" #Insert model name here
cmod1 <- h2o.loadModel(paste0("saved-models/", modelname1))
```

Variable importance:
```{r}
vi1 <- h2o.varimp(cmod1)
vi1
h2o.varimp_plot(cmod1, num_of_features = length(X1))
```

Model Performance:
```{r}
set.seed(2025)
pred <- h2o.predict(cmod1, newdata = data_test)
(perf <- h2o.performance(cmod1, newdata = data_test, valid=T))
```

Model Performance with another packages:
```{r}
data_test_r <- as.data.frame(data_test)
pred_r <- as.data.frame(pred)

roc <- roc(data_test_r$is_goal, pred_r$p1)
plot.roc(roc, print.auc = TRUE, print.thres = "best")

best_thres <- coords(roc, "best", ret = "threshold", transpose = T)
h2o.confusionMatrix(cmod1, data_test, thresholds = best_thres, metrics = "mean_per_class_accuracy")

pred_r$xG_bool <- ifelse(pred_r$p1 > best_thres, 1, 0)
(result <- confusionMatrix(data = factor(pred_r$xG_bool), 
                           reference = factor(data_test_r$is_goal), positive = "1", mode="prec_recall"))
```

Loading GLM model with only 2 features:
```{r}
GLM_modelname1 <- "" #Insert model name here
GLM_cmod1 <- h2o.loadModel(paste0("saved-models/", GLM_modelname1))
```


GLM Variable importance:
```{r}
GLM_vi1 <- h2o.varimp(GLM_cmod1)
GLM_vi1
h2o.varimp_plot(GLM_cmod1, num_of_features = length(X1))
```

GLM Model Performance:
```{r}
set.seed(2025)
GLM_pred <- h2o.predict(GLM_cmod1, newdata = data_test)
(GLM_perf <- h2o.performance(GLM_cmod1, newdata = data_test, valid=T))
```

GLM Model Performance with another packages:
```{r}
GLM_data_test_r <- as.data.frame(data_test)
GLM_pred_r <- as.data.frame(GLM_pred)

GLM_roc <- roc(GLM_data_test_r$is_goal, GLM_pred_r$p1)
plot.roc(GLM_roc, print.auc = TRUE, print.thres = "best")

GLM_best_thres <- coords(GLM_roc, "best", ret = "threshold", transpose = T)
h2o.confusionMatrix(GLM_cmod1, data_test, thresholds = GLM_best_thres, metrics = "mean_per_class_accuracy")

GLM_pred_r$xG_bool <- ifelse(GLM_pred_r$p1 > GLM_best_thres, 1, 0)
(GLM_result <- confusionMatrix(data = factor(GLM_pred_r$xG_bool), 
                           reference = factor(GLM_data_test_r$is_goal), positive = "1", mode="prec_recall"))
```

Results analysis:
```{r}
set.seed(2025)
pred_1 <- h2o.predict(cmod1, newdata = data_test)
pred_r_1 <- as.data.frame(pred_1)

pred_2 <- h2o.predict(cmod2, newdata = data_test)
pred_r_2 <- as.data.frame(pred_2)
```

xG values distribution:
```{r}
ggplot(pred_r_2, aes(x=p1)) + 
        geom_density() +
        labs(x="xG") +
        theme_bw() +
        scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
        scale_y_continuous(limits = c(0, 8), expand = c(0, 0))

ggsave("density_xg_plot.png", width = 10, height = 5)

data_test_r <- as.data.frame(data_test)
xx <- data_test_r %>% mutate(xG = pred_r_2$p1)

ggplot(data = xx, aes(x= x1, y = y1)) + 
        annotate_pitch(colour = "white",
                       fill   = "black",
                       limits = FALSE) +
        theme_pitch() +
        theme(plot.background = element_rect(fill = "black"),
              title = element_text(colour = "white")) +
        coord_flip(xlim = c(50, 100),
                   ylim = c(0, 100)) +
        geom_tile(aes(fill = xG)) +
        scale_fill_gradientn(colours = rev(brewer.pal(7, "Spectral"))) +
        theme(legend.position = c(0.76, 1), legend.direction = "horizontal",
              legend.text = element_text(color = "white", size = 8, face = "plain"),
              legend.background = element_rect(fill = "black"),
              legend.key = element_rect(fill = "black", color = "black"),
              plot.title = element_text(hjust = 0.07, face = "plain"),
              plot.subtitle = element_text(hjust = 0.07, size = 10, face = "italic"),
              plot.caption = element_text(hjust = 0.95),
              plot.margin = margin(1, 0.2, 0.5, 0.2, "cm")) +
        labs(fill = "xG value")


ggsave("xG_plot2.png", width = 10*1.3, height = 5*1.6)
```

ROC curve:
```{r}
preds_list <- list(pred_r_1$p1, pred_r_2$p1)

nm <- length(preds_list)
goal <- as.data.frame(as.numeric(as.character(data_test$is_goal)))
actuals_list <- rep(goal, nm)

pred <- prediction(preds_list, actuals_list)
rocs <- performance(pred, "tpr", "fpr")

png(filename = "AUC_plot.png", width = 8, height = 5, units = "in", res = 300)
plot(rocs, col = as.list(1:nm), main = "Testing dataset ROC Curves")
legend(x = "bottomright", 
       legend = c("GBM 2", "GBM all"),
       fill = 1:nm)
dev.off()
```

Partial dependency plots:
```{r}
set.seed(2025)
custom_predict <- function(model, newdata) {
        newdata_h2o <- as.h2o(newdata)
        res <- as.data.frame(h2o.predict(model, newdata_h2o))
        return(as.numeric(as.character(res$predict)))
}

explainer_gbm_1 <- DALEX::explain(model = cmod1, 
                                    data = as.data.frame(data_test)[, X1],
                                    y = as.numeric(as.character(as.data.frame(data_test)[, Y])),
                                    predict_function = custom_predict,
                                    label = "GBM 2")
mp_gbm_1 <- model_performance(explainer_gbm_1)

explainer_gbm_2 <- DALEX::explain(model = cmod2, 
                                    data = as.data.frame(data_test)[, X2],
                                    y = as.numeric(as.character(as.data.frame(data_test)[, Y])),
                                    predict_function = custom_predict,
                                    label = "GBM all")
mp_gbm_2 <- model_performance(explainer_gbm_2)
```

```{r}
set.seed(2025)
pdp_gbm_1 <- partial_dependency(explainer_gbm_1, variables = "distance_to_goal_line")
pdp_gbm_2 <- partial_dependency(explainer_gbm_2, variables = "distance_to_goal_line")
```
Plot with GG plot:
```{r}
set.seed(2025)
ggplot() +
  geom_smooth(data = pdp_gbm_1, aes(x = `_x_`, y = `_yhat_`, color = "GBM 2"), 
              se = FALSE, method = "loess", size = 1) +
  geom_smooth(data = pdp_gbm_2, aes(x = `_x_`, y = `_yhat_`, color = "GBM all"), 
              se = FALSE, method = "loess", size = 1) +
  scale_x_continuous(breaks = seq(0, 30, by = 5), limits = c(0, 30)) +
  scale_y_continuous(limits = c(0, 1))+  # Added y-axis limits
  labs(x = "Distance to Goal Line (m)", 
       y = "Average Prediction",
       color = "Model") +
  theme_bw()

ggsave("Partial_distance.png", width = 8, height = 5)
```


Rate of decline:
```{r}
set.seed(2025)
# Calculate the rate of change (derivative) for each model
calculate_rate <- function(pdp_data) {
  pdp_data %>%
    arrange(`_x_`) %>%
    mutate(
      xG_change = c(NA, diff(`_yhat_`)),
      distance_change = c(NA, diff(`_x_`)),
      rate_of_decline = xG_change / distance_change
    )
}

gbm1_rate <- calculate_rate(pdp_gbm_1) %>% mutate(model = "GBM 2")
gbm2_rate <- calculate_rate(pdp_gbm_2) %>% mutate(model = "GBM all")

# Combine data
combined_rates <- bind_rows(gbm1_rate, gbm2_rate)
```

Plot the rate of decline:
```{r}
set.seed(2025)
ggplot(combined_rates, aes(x = `_x_`, y = rate_of_decline, color = model)) +
  geom_smooth(se = FALSE, method = "loess", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_x_continuous(breaks = seq(0, 30, by = 5), limits = c(0, 30)) +
  scale_color_manual(values = c("GBM 2" = "#F8766D", "GBM all" = "#00BFC4")) +
  labs(
    x = "Distance to Goal Line (m)",
    y = "Rate of xG Decline (per meter)",
    color = "Model",
    title = "How quickly does xG decrease with distance?",
    subtitle = "Negative values = xG declining, steeper = faster decline"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, color = "gray30")
  )
ggsave("xG_decline_rate.png", width = 10, height = 6)
```


Summary:
```{r}
set.seed(2025)
summary_stats <- combined_rates %>%
  filter(`_x_` <= 40, !is.na(rate_of_decline)) %>%
  group_by(model) %>%
  summarise(
    avg_decline_rate = mean(rate_of_decline, na.rm = TRUE),
    steepest_decline = min(rate_of_decline, na.rm = TRUE),
    distance_at_steepest = `_x_`[which.min(rate_of_decline)]
  )

print(summary_stats)
```


Plot angle to goal with GGplot:
```{r}
set.seed(2025)
pdp_gbm_1_angle <- partial_dependency(explainer_gbm_1, variables = "angle_to_goal")
pdp_gbm_2_angle <- partial_dependency(explainer_gbm_2, variables = "angle_to_goal")

ggplot() +
  geom_smooth(data = pdp_gbm_1_angle, aes(x = `_x_`, y = `_yhat_`, color = "GBM 2"), 
              se = FALSE, method = "loess", size = 1) +
  geom_smooth(data = pdp_gbm_2_angle, aes(x = `_x_`, y = `_yhat_`, color = "GBM all"), 
              se = FALSE, method = "loess", size = 1) +
  scale_x_continuous(breaks = seq(0, 90, by = 5), limits = c(0, 90)) +
  scale_y_continuous(limits = c(0, 1))+  # Customize breaks here
  labs(x = "Angle to Goal (degrees)", 
       y = "Average Prediction",
       color = "Model") +
  theme_bw()

ggsave("Partial_angle.png", width = 8, height = 5)
```



Skilled foot plot:
```{r}
pdp_gbm_2_foot <- partial_dependency(explainer_gbm_2, variables = "skilled_foot")
plot(pdp_gbm_2_foot)
```
Save the plot:
```{r}
png(filename = "Partial_foot.png", width = 8, height = 5, units = "in", res = 300)
plot(pdp_gbm_2_foot)
dev.off()
```




GLM Results analysis:
```{r}
set.seed(2025)
GLM_pred_1 <- h2o.predict(GLM_cmod1, newdata = data_test)
GLM_pred_r_1 <- as.data.frame(GLM_pred_1)

GLM_pred_2 <- h2o.predict(GLM_cmod2, newdata = data_test)
GLM_pred_r_2 <- as.data.frame(GLM_pred_2)
```

xG values distribution:
```{r}
ggplot(GLM_pred_r_2, aes(x=p1)) + 
        geom_density() +
        labs(x="xG") +
        theme_bw() +
        scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
        scale_y_continuous(limits = c(0, 8), expand = c(0, 0))

ggsave("GLM_density_xg_plot.png", width = 10, height = 5)

GLM_data_test_r <- as.data.frame(data_test)
GLM_xx <- GLM_data_test_r %>% mutate(xG = GLM_pred_r_2$p1)

ggplot(data = GLM_xx, aes(x= x1, y = y1)) + 
        annotate_pitch(colour = "white",
                       fill   = "black",
                       limits = FALSE) +
        theme_pitch() +
        theme(plot.background = element_rect(fill = "black"),
              title = element_text(colour = "white")) +
        coord_flip(xlim = c(50, 100),
                   ylim = c(0, 100)) +
        geom_tile(aes(fill = xG)) +
        scale_fill_gradientn(colours = rev(brewer.pal(7, "Spectral"))) +
        theme(legend.position = c(0.76, 1), legend.direction = "horizontal",
              legend.text = element_text(color = "white", size = 8, face = "plain"),
              legend.background = element_rect(fill = "black"),
              legend.key = element_rect(fill = "black", color = "black"),
              plot.title = element_text(hjust = 0.07, face = "plain"),
              plot.subtitle = element_text(hjust = 0.07, size = 10, face = "italic"),
              plot.caption = element_text(hjust = 0.95),
              plot.margin = margin(1, 0.2, 0.5, 0.2, "cm")) +
        labs(fill = "xG value")


ggsave("GLM_xG_plot2.png", width = 10*1.3, height = 5*1.6)
```


GLM ROC curve:
```{r}
set.seed(2025)
GLM_preds_list <- list(GLM_pred_r_1$p1, GLM_pred_r_2$p1)

nm <- length(GLM_preds_list)
goal <- as.data.frame(as.numeric(as.character(data_test$is_goal)))
actuals_list <- rep(goal, nm)

GLM_pred <- prediction(GLM_preds_list, actuals_list)
GLM_rocs <- performance(GLM_pred, "tpr", "fpr")

png(filename = "GLM_AUC_plot.png", width = 8, height = 5, units = "in", res = 300)
plot(rocs, col = as.list(1:nm), main = "Testing dataset ROC Curves")
legend(x = "bottomright", 
       legend = c("GLM 2", "GLM all"),
       fill = 1:nm)
dev.off()
```

Partial dependency plots:
```{r}
set.seed(2025)
custom_predict <- function(model, newdata) {
        newdata_h2o <- as.h2o(newdata)
        res <- as.data.frame(h2o.predict(model, newdata_h2o))
        return(as.numeric(as.character(res$predict)))
}

explainer_glm_1 <- DALEX::explain(model = GLM_cmod1, 
                                    data = as.data.frame(data_test)[, X1],
                                    y = as.numeric(as.character(as.data.frame(data_test)[, Y])),
                                    predict_function = custom_predict,
                                    label = "GLM 2")
mp_glm_1 <- model_performance(explainer_glm_1)

explainer_glm_2 <- DALEX::explain(model = GLM_cmod2, 
                                    data = as.data.frame(data_test)[, X2],
                                    y = as.numeric(as.character(as.data.frame(data_test)[, Y])),
                                    predict_function = custom_predict,
                                    label = "GLM all")
mp_glm_2 <- model_performance(explainer_glm_2)
```
Same approach for GLM models:

```{r}
pdp_glm_1 <- partial_dependency(explainer_glm_1, variables = "distance_to_goal_line")
pdp_glm_2 <- partial_dependency(explainer_glm_2, variables = "distance_to_goal_line")

ggplot() +
  geom_smooth(data = pdp_glm_1, aes(x = `_x_`, y = `_yhat_`, color = "GLM 2"), 
              se = FALSE, method = "loess", size = 1) +
  geom_smooth(data = pdp_glm_2, aes(x = `_x_`, y = `_yhat_`, color = "GLM all"), 
              se = FALSE, method = "loess", size = 1) +
  scale_x_continuous(breaks = seq(0, 30, by = 5), limits = c(0, 30)) +
  scale_y_continuous(limits = c(0, 1))+
  labs(x = "Distance to Goal Line (m)", 
       y = "Average Prediction",
       color = "Model") +
  theme_bw()

ggsave("GLM_Partial_distance.png", width = 8, height = 5)
```

Rate of decline
```{r}
set.seed(2025)
glm1_rate <- calculate_rate(pdp_glm_1) %>% mutate(model = "GLM 2")
glm2_rate <- calculate_rate(pdp_glm_2) %>% mutate(model = "GLM all")

# Combine data
combined_rates <- bind_rows(glm1_rate, glm2_rate)
```

Plot the rate of decline:
```{r}
ggplot(combined_rates, aes(x = `_x_`, y = rate_of_decline, color = model)) +
  geom_smooth(se = FALSE, method = "loess", size = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_x_continuous(breaks = seq(0, 30, by = 5), limits = c(0, 30)) +
  scale_color_manual(values = c("GLM 2" = "#F8766D", "GLM all" = "#00BFC4")) +
  labs(
    x = "Distance to Goal Line (m)",
    y = "Rate of xG Decline (per meter)",
    color = "Model",
    title = "How quickly does xG decrease with distance?",
    subtitle = "Negative values = xG declining, steeper = faster decline"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, color = "gray30")
  )
ggsave("xG_decline_rate_GLM.png", width = 10, height = 6)

```


Summary:
```{r}
set.seed(2025)
summary_stats <- combined_rates %>%
  filter(`_x_` <= 40, !is.na(rate_of_decline)) %>%
  group_by(model) %>%
  summarise(
    avg_decline_rate = mean(rate_of_decline, na.rm = TRUE),
    steepest_decline = min(rate_of_decline, na.rm = TRUE),
    distance_at_steepest = `_x_`[which.min(rate_of_decline)]
  )

print(summary_stats)
```


```{r}
set.seed(2025)
pdp_glm_1_angle <- partial_dependency(explainer_glm_1, variables = "angle_to_goal")
pdp_glm_2_angle <- partial_dependency(explainer_glm_2, variables = "angle_to_goal")

ggplot() +
  geom_smooth(data = pdp_glm_1_angle, aes(x = `_x_`, y = `_yhat_`, color = "GLM 2"), 
              se = FALSE, method = "loess", size = 1) +
  geom_smooth(data = pdp_glm_2_angle, aes(x = `_x_`, y = `_yhat_`, color = "GLM all"), 
              se = FALSE, method = "loess", size = 1) +
  scale_x_continuous(breaks = seq(0, 90, by = 5), limits = c(0, 90)) +
  scale_y_continuous(limits = c(0, 1))+  # Customize breaks here
  labs(x = "Angle to Goal (degrees)", 
       y = "Average Prediction",
       color = "Model") +
  theme_bw()

ggsave("GLM_Partial_angle.png", width = 8, height = 5)
```




```{r}
set.seed(2025)
pdp_glm_2_foot <- partial_dependency(explainer_glm_2, variables = "skilled_foot")
plot(pdp_glm_2_foot)
```


```{r}
png(filename = "GLM_Partial_foot.png", width = 8, height = 5, units = "in", res = 300)
plot(pdp_glm_2_foot)
dev.off()
```







